---
title       : Calculating Proton Beam Effects For Cancer Therapy
subtitle    : 
author      : Seth A. Veitzer
job         : 
# io2012 does not support italics or bold text. Reason enough.
#framework   : html5slides # {io2012, html5slides, shower, dzslides, ...}
framework   : io2012 # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
--- 
<!-- Copyright &copy; 2015 Seth A. Veitzer                                  -->
<!-- Please do not plagarize                                                -->

## Knowing Proton Stopping Powers Enables Effective Cancer Treatment

* Proton beam therapy treats malignant tumors in cases where
traditional radiation therapy is not possible, e.g. for ocular tumors (eye)
* Protons will deposit most of their energy in a much smaller volume than
radiation, minimizing damage to healthy tissue
* The rate at which protons deposit energy in a target material is called
*Stopping Power*, measured in using of $MeV \cdot cm^2/g$
* Proton stopping power depends on both target material properties and 
on the proton energy
  - Total stopping power has a peak value (Bragg Peak), where the rate of 
energy loss is greatest
  - Most of a proton's energy (>90%) will be deposited at the Bragg Peak

--- &twoColumn

## The Stopping Power App Makes It Easy To Compare Stopping Powers

*** =left

* The Stopping Power WebApp reads in stopping power data for up to six
different materials at a time over a large range
of incident proton energies using R
* The app plots proton stopping powers for different materials when they
are selected on the side panel
  - Users can directly compare different materials over a large range of proton 
energies, measured in $MeV$
  - Plots can be made on a linear or log-log scale, depending on the energy range
of interest

*** =right

``` {r echo=FALSE}
stopping <- read.csv('stopping.csv', sep=' ')
xlim <- c(0,2); ylim <- c(0,4000)

plot(stopping[,1],stopping[,1], type='n', xlim=xlim, ylim=ylim,
  xlab='Proton Kinetic Energy (MeV)',
  ylab='Stopping Power (Mev cm^2/g)', main='Total Stopping Power')

lines(stopping[,1],stopping[,2], lwd=3, col='red')
lines(stopping[,1],stopping[,6], lwd=3, col='blue')
legend <- c('Hydrogen', 'Striated Muscle'); legendCols <- c('red', 'blue')
legend('topright', legend, lty=1, lwd=3, col=legendCols)
```

--- &twoColumn

## Average Stopping Power Can Be Computed

*** =left

* The Stopping Power WebApp provides a button on the side panel will calculate and display the
mean value of proton stopping power
  - This demonstrates reactive output based on the UI input and real-time server calculations
  - Only stopping powers that are currently being plotted are included in the average
  - Provides a quick way to estimate the actual stopping power of materials that have
many different compositions

*** =right

``` {r echo=FALSE}
stopping <- read.csv('stopping.csv', sep=' ')
xlim <- c(0,2); ylim <- c(0,4000)

plot(stopping[,1],stopping[,1], type='n', xlim=xlim, ylim=ylim,
  xlab='Proton Kinetic Energy (MeV)',
  ylab='Stopping Power (Mev cm^2/g)', main='Total Stopping Power')

lines(stopping[,1],stopping[,2], lwd=3, col='red')
lines(stopping[,1],stopping[,6], lwd=3, col='blue')
lines(stopping[,1],(stopping[,2]+stopping[,6])/2., lwd=3, col='black')
legend <- c('Hydrogen', 'Striated Muscle', 'Mean Stopping Power'); 
legendCols <- c('red', 'blue','black')
legend('topright', legend, lty=1, lwd=3, col=legendCols)
```

---

## Conclusions

* The Stopping Power WebApp can be used to plot and compare proton stopping powers
for different target materials
* This tool can aid in the tuning of proton beam energy for cancer therapy, based on the
types of tissue and the depth of the tumor to be treated
* Average stopping power can be computed and displayed to model materials that
are composed of multiple different types of materials
* The data used in this WebApp was scraped from the NIST website
http://www.nist.gov/pml/data/star/index.cfm that publishes tables
of stopping powers for electrons, protons, and helium atoms (alpha particles)


